Towards a Machine Learning approach for Aggregate Computing

### 2021-11-07 paper-2021-woa-ac-ml.tex
refinements to whole document: grammar and content

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
  T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\sloppy

\newcommand{\unibo}{\textsc{Alma Mater Studiorum} -- \textit{Università di Bologna}}
\newcommand{\disi}{\textit{Computer Science and Engineering - DISI}}
\newcommand{\scafi}{{ScaFi}}
\newcommand{\scalaloci}{{ScalaLoci}}
\newcommand{\scafiloci}{{ScaFiLoci}}
\newcommand{\scalainline}[1]{\mintinline[fontsize=\small]{scala}{#1}}
%%%%% LEGENDA FOR NOTES %%%%%
%% RED: constraints
%% YELLOW: suggestions
%% CYAN: todos
%% GREEN: outline
\newcommand{\constraints}[1]{\todo[inline, color=red]{#1}}
\newcommand{\suggestions}[1]{\todo[inline, color=yellow]{#1}}
\newcommand{\todos}[1]{\todo[inline, color=cyan]{\textbf{TODO}: #1}}
\newcommand{\outline}[1]{\todo[inline, color=green]{#1}}
%% Utility commands
\newcommand{\hybridaggregate}{\textit{Hybrid Aggregate Computing}}
\newcommand{\round}{\texttt{round}}
\newcommand{\export}{\texttt{export}}

\begin{document}

@@

One of the most challenging fields for Machine Learning application is Collective-Adaptive System (CAS). % Or do we want to use Cyber-Physical Swarm??
 Here, engineers have to handle large scale, global-to-local behaviour mapping,
 highly environment stochasticity. 
%
Historically, researchers favour handling them with self-organisation,
 by which a collective behaviour emerges from continuously local interaction between simple components.
%
A novel technique to specify self-organising behaviour in a composable and functional manner is Aggregate Computing.
 In particular, this paradigm enables collective program specification as a whole, 
 as it is based on a distributed data structure manipulation called \textit{computational fields}.
 It is successfully used in different scenarios ranging from Swarm Robotics to Smart cities.

Even though it is promising for expressing high-level collective behaviour system-size independent,
 practically it is difficult to make programs work in highly dynamic networks, 
 where nodes have to deal with continuous changes at the environment side.
%
Therefore, with this work, we stand the first step towards integration between Aggregate Computing and Machine Learning,
 in order to improve adaptivity acting in the low-level specifications but maintaining the high-level program behaviours.

@@

\section{Introduction}
Nowadays, computer engineers have to deal with Collective Adaptive Systems (CASs)
 due to the increasing interconnected computational devices placed in ever-changing environments.
 Examples of those include robot swarm, smart cities and augmented crowds with smart devices.

CAS are notoriously hard~\cite{DBLP:conf/huc/Ferscha15} to design as they are characterised by 
\begin{itemize}
\item a global order that emerges from local node interactions, 
\item decentralised control, and 
\item adaptive behaviours towards environmental changes to pursuit a collective goal.
\end{itemize} 
%
So, over the years, in literature, various approaches aim at handling this complexity. 
%
Some of them take inspiration from the natural system
 composed of social animals --- like bees, fish and ants. 
%
Here, a sort of hive mind arises -- 
 referred also as swarm intelligence -- 
born from local interactions between agents. 
%
These kinds of behaviours are accomplished with self-organisation, 
 namely a global order that emerges from continuous 
 interactions between simple entities.

Initially, researchers try to bring those properties into the computer science world 
 by miming the collective behaviour bottom-up~\cite{DBLP:journals/connection/Webb02}, 
 namely trying to find the right single node behaviour 
 to reach a global behaviour, e.g. flocking, foraging, etc.
%
Though, this mapping isn't easily definable ---
 since we handle complex systems.
%
Ideally, we would like to express collective behaviours focusing only at the ensemble level,
 abstracting over non-functional aspects such as the network topology, environment uncertainties and 
 system size.

In this direction, Aggregate Computing~\cite{DBLP:journals/computer/BealPV15} is an innovative solution by which 
 developers can express collective self-organising behaviours through a functional program specification.
%
Indeed, an aggregate program (i.e. a program written with Aggregate Computing) consists of the manipulation of \textit{computational fields}, a distributed
 data structure where each point in the spacetime is associated with the result of a computation
 performed by a node.
%
These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
% This calculus describes a set of minimal constructs to express any spatiotemporal computation.

In Aggregate Computing there are common abstractions in collective behaviour specification. 
 These are encapsulated in so-called ``building blocks''.
%
With those, different properties have been proven (such as \textit{eventual consistency}~\cite{DBLP:journals/taas/BealVPD17}) and various
 applications have been built (such as swarm robotics, crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).

Though, building block synthesis concerns different stuff related not only to the behaviour itself
 but also to the "ensemble" specifications. 
 For instance, it is quite difficult to craft building blocks
 that easily adapt w.r.t the node movements rapidity.

Ideally, we would continue to build our application using these building blocks -- so the high-level specification remains the same -- using some
 under-the-hood mechanisms that adapt the behaviour according to different environmental changes.

@@

In particular, In this paper,
 we investigate how Aggregate Computing can be extended with Reinforcement Learning (RL) through a case study,
 where we outline the performance improvement of the "hop count" program. 
 Furthermore, we show that the learnt policy scales up to different system sizes.
\todos{Outline}

@@

In particular, in literature, we speak of Collective Adaptive Systems
 when the number is incredibly high, there is no central authority that coordinate nodes,
 the overall behaviour is reached with local interactions, the systems expose self-* properties,
 and the ensemble pursuit of collective goals
 exhibiting swarm intelligence.

Engineering this kind of system is a thriving area, various approaches have been proposed.
 First attempts leverage so-called \textit{local-to-global} techniques, where the collective behaviour emerges from local behaviour description, 
 usually inspired by nature~\cite{DBLP:journals/swarm/BrambillaFBD13}. 
%
Though, a natural way to describe this kind of problem is by using an \textit{collective} view of the system.
 And it is the idea used by the \textit{global-to-local} techniques~\cite{DBLP:journals/jlap/ViroliBDACP19,DBLP:journals/scp/AlrahmanNL20, DBLP:conf/cbse/BuresGHKKP13}, where the behaviours are designed \textit{top-down},
 and then the local behaviour is generated automatically. 
% 
Also ``meet-in-the-middle''~\cite{DBLP:journals/computer/PinciroliB16} approaches exist, where
 a part of the system is still expressed bottom-up (device-centric) whereas a part is expressed top-down (swarm-centric)

In our works, we are focused on Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}, global-to-local approach where
 the self-organising collective behaviour is expressed in terms of manipulation of \textit{computational fields}.

The latter could be seen as the extension of electromagnetic fields to computer science since it basically consists in a 
 map from any device to its computational value (i.e. the result of a computation).

The computational field manipulation is ruled by \textit{field calculus}, which define the constructs needed 
 any for express spatio-temporal computation~\cite{DBLP:conf/coordination/AudritoBDV18}.

The novelty of Aggregate Computing consists in its composable (functional-like) program definition: using basic constructs
 we identify the main patterns (\textit{building-blocks}) that could be reusable at the application level.

Crafting opportunistically these building-blocks is possible to verify prominent properties such as eventual consistency~\cite{DBLP:journals/taas/BealVPD17} (i.e. the ensamble behaviour is independent from the underlying network details)
 and self-stabilisation~\cite{DBLP:journals/tomacs/ViroliABDP18} (i.e. the ability of a system to recover from arbitrary changes).

Aggregate Computing is studied in various simulated scenarios (thanks to Alchemist, a meta-simulator for pervasive systems), ranging from smart cities to crowds, thanks to programming language and toolchain that support 
 this paradigm, such as ScaFi~\cite{DBLP:conf/ecoop/CasadeiV16}, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.

In the next sections, we show the field calculus operators that are essential to understand the case study presented,
 and then we underline how each node computes in order to follow the aggregate program specification.
\subsection{Field Calculus operators}
The main constructs that capture the essential aspects for programming self-organising systems are:
\begin{itemize}
  \item \textit{Stateful field evolution} expression \mintinline[escapeinside=||]{scala}{|{\color{teal}rep}|(|$e_1$|) {|($x$) => $e_2$|}} describes a field evolving in time. 
  $e_1$ is the initial field value and the function $(x) => e_2$ define how the field changes at each execution.
  \item \textit{Neighbour interaction} expression \mintinline[escapeinside=||]{scala}{|\color{teal}nbr|{|$e$|}} a view
  of the field values in the surroundings of each device where neighbours are
  mapped to their evaluations of $e$. A series of \mintinline{scala}{*hood} operators leverage this map to produce relevant result. 
  For instance, $minHood$ return the minimum value 
  from the neighbourhoods.
  \item \textit{Domain partitioning} expression \mintinline[escapeinside=||]{scala}{|\color{teal}if|(|$e_0$|){|$e_1$|}{|$e_2$|}} splits the computational
  field into two non-communicating zones hosting isolated subcomputations:
  $e_1$ where $e_0$ is true, and $e_2$ where $e_0$ is false.
\end{itemize}

To a more in depth explanation of these opearatos, please refear to~\cite{DBLP:journals/computer/BealPV15}.
\subsection{Computational Model}
An aggregate program execution consists of a continuously node-point execution of a \round{}, 
 that is the atomical part of aggregate computing.
%
Each node can interact only with-in its neighbourhoods sharing the \export{}, 
 namely the result produced by the evaluation of a \round{}.

The main round steps consist in:
\begin{enumerate}
  \item execution context creation: where node collects information from the sensors, the neighbourhoods exports and the old
  program output
  \item program evaluation: the system-wide aggregate program is evaluated against the local context created by the nodes. 
  This produces the \export{}.
  \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes
  \item actuations: giving the export produced, nodes may perform some actuation (e.g. motor activations, steering, ...)
\end{enumerate}

@@

In this section,
 we analyse the characteristics of the aggregate computing paradigm,
 map these to relevant machine learning contributions in literature,
 pointing out different learning perspectives about the integration of Aggregate Computing with Machine Learning,
 and summarise the analysis into a roadmap.

\paragraph{Multi-and many-agent system}
%
An aggregate system is a multi-agent system
 and, often, a \emph{many}-agent system
 where a large number (hundreds or more)
 of autonomous entities are programmed to achieve 
 some collective behaviour by means of \emph{repeated} 
 sensing, computation, communication, and actuation steps.
%
Due to the high stochasticity of the environment,
 it is almost impossible to know and
 program the optimal behaviour for all agents in advance.
 This results in the need to create intelligent agents
 so that they can learn the optimal behaviour and adapt to environmental changes.
%
In recent decades there has been an emerging trend in the use of Reinforcement Learning 
 in multi-agent settings, as a powerful, robust and adaptive learning paradigm.
 Progress has been considerable and a wide range of algorithms are now available.
% to expand and references
This field is a conjunction of Game theory and Reinforcement Learning, 
 and there are several (even orthogonal) viewpoints on which researchers have been focused.

First attempts from RL viewpoint, 
 goes towards a so-called \textit{conccurent/independent learning}~\cite{DBLP:journals/tsmc/BusoniuBS08} approach, 
 where each agent learn locally against the whole environment~\cite{DBLP:conf/icml/Tan93}. 
Here various approaches have been proposed,
 such as Hysteretic Q-Learning (used to reduce non-stationarity)~\cite{hysteretic-q}, 
 Lenient Learning~\cite{DBLP:journals/jmlr/WeiL16}, and Frequency Maximum Q-value.
The pro of these approaches is that their complexity does not scale up with the number of agents, 
 but, unfortunately, the learning process is extremely non-stationary and non-stable.
 Furthermore, they are heuristic and does not exist any convergence proof to a stationary state (even if practically often convergence to a stable policy).

%
The game theory viewpoint instead has tried to find techniques that achieve equilibrium such as Nash-equilibrium of Pareto optimality.
 The first works towards this direction are Nash-Q-Learning~\cite{nash-q}, Friend-or-Foe Q-learning~\cite{DBLP:conf/icml/Littman01}, Minimax-Q.
The problem there is that is used with few agents and scale hardly with the application complexity.
%
So, current emerging trends tend to leverage a so-called centralised training and decentralised execution (CTDE) approach, by which 
 agents should leverage the system-wide knowledge at training time but at runtime, they act independently~\cite{DBLP:journals/aamas/Hernandez-LealK19}.
 Most promising approaches in this direction are MAPPO~\cite{DBLP:journals/corr/abs-2103-01955}, COMA~\cite{DBLP:journals/corr/FoersterFANW17} and MADDPG~\cite{DBLP:conf/nips/LoweWTHAM17}.

%
 The main problem with most of the solutions available in the literature is that they consider a small number of agents (or at least test them on small games).

\paragraph{Neighbour-based or indirect (environment-mediated) interaction}
%
In aggregate systems, a device can only directly interact with neighbours.
%
Data flows may be implemented
 to support indirect communication across multiple hops;
 however, information from devices far away tends to be obsolete.
%
The environment can also be used, via stigmergy,
 for indirect communication;
 however, a device can typically only access 
  to a very local portion of the environment.
%
In other words, the system state is only \emph{partially observable}.

@@

A way to handle this kind of rewards in multi-agent settings is \textit{difference rewards}~\cite{DBLP:conf/atal/DevlinYKT14}. It is used to capture agents contributions to the system's global performance.
\paragraph{Transient phase}
%
When an aggregate behaviour is carried out,
 we often focus on the stabilised outcomes,
 tolerating outcomes that are being generated
 during the transient phase.
%
However, intermediate results are often important,
 and invariants should hold throughout the entire computation.

@@

\subsection{Integration perspectives}
Here we outline the different perspectives about Aggregate Computing with learning, for
 of clarifying the current and the future research path.
%
\paragraph{Application} Learning can be useful at different levels in our framework.
 First of all, learning can be applied to the building blocks definition.
% 
Indeed, most of our results and applications are based on them.
%
The building block improvement may consist either of learning a correction factor or by learning the behaviour entirely.

On the other side, learning can be useful also for framework related stuff,
 mainly aim at "non-functional" aspects. 
 Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that currently have been defined by hand.

\paragraph{Learning problem} 
The learning problem in multi-agent settings can be pose as \textit{Independent} and \textit{Team} learning. 
In the foster, each agent perform the learning process and in the latter,
 the learning process is performed only by one agent and then shared with the collective.
% 
Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour 
 --- i.e. each agent performs the same program. 
 Here, a common reward signal exists and evaluate the overall collective behaviour.
% 
In Aggregate Computing, each agent follows the \textit{same} aggregate program shared within the entire system, so Team learning can 
 be applied seamlessly.

On the other hand, Independent Learning is usually applied in Collective
 (Self-)Adaptive System \cite{csas-and-marl}, a field near to Aggregate Computing.
 In particular, Independent Reinforcement Learning reach good results in this kind of system, 
 even if no theorems exist as in the single-agent context.
%
With Independent learning though is still possible to reach a global collective behaviour convergence. 
 Indeed using a common reward signal, the collective may reach the same behaviour through emergence \cite{iima2008swarm, nguyen2018swarm}.

\paragraph{Learning technique}
Aside from what kind of problem we set up, we can then use different Machine Learning techniques,
 such as Reinforcement Learning or Supervised Learning.
%
Supervised learning is rare in CSAS -- or in the ``swarm" like system in general.
%
Novel works~\cite{DBLP:conf/corl/TolstayaGPP0R19} leverage Graph Neural network to learn how agents should communicate, 
 supposing to know the right behaviour (i.e. produced via global vision simulation or by emulation of natural phenomena).
%
In Aggregate Computing, Supervised Learning can be applied when we want to craft new building blocks by zero. 
% 
Indeed, we know eventually the right state of the system, but we do not know how the system can reach it.

RL is predominantly used in CAS, 
 due to its flexibility and thanks to the plethora of approaches used in this scenario~\cite{csas-and-marl}.
% 
In our case, RL can be used both at the building block level and the framework level.
%
For the foster, the simplest scenario can be that agents learn a correction factor to improve a long-term reward signal.

\paragraph{Learning time}
Finally, learning can be performed \textit{offline}, \texttt{online} or \textit{offine an then online} (\textit{mixed time}).
% 
The foster is the common scenario in multi-agent settings.
 Here indeed, we can leverage simulation to know the right computational field and then use one of the novel CTDE techniques.

Mixed-time learning can be a mid-level complexity direction instead.
 Here, simulation can always be leveraged to reach a good behaviour, but at runtime, the behaviour can be adapted in front of new environmental situations.

@@

\section{Aggregate Computing with Reinforcement}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{img/algorithm-learning.pdf}
  \caption{This figure shows the main steps of RL in Aggregate Computing.
  Each agent starts with his copy of a q-table. 
  Then, during the simulation, they refine them q-tables changing q-values according to their experience (different colours mean different values). 
  In the end, a new $Q_{max}$ table will be created aggregating each agent table. 
  Then, a new episode starts using this $Q_{max}$ as a starting point.
  }
  \label{fig:aggregate-q-learning}
\end{figure}
As first experiments, we leverage RL to improve building blocks. 
% 
In this section, we briefly describe how Reinforcement Learning can be injected into the computation loop,
 and what Reinforcement Learning we used with Aggregate Computing.

>>>>

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
  T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\newcommand{\unibo}{\textsc{Alma Mater Studiorum} -- \textit{Università di Bologna}}
\newcommand{\disi}{\textit{Computer Science and Engineering - DISI}}
\newcommand{\scafi}{{ScaFi}}
\newcommand{\scalaloci}{{ScalaLoci}}
\newcommand{\scafiloci}{{ScaFiLoci}}
\newcommand{\scalainline}[1]{\mintinline[fontsize=\small]{scala}{#1}}
%%%%% LEGENDA FOR NOTES %%%%%
%% RED: constraints
%% YELLOW: suggestions
%% CYAN: todos
%% GREEN: outline
\newcommand{\constraints}[1]{\todo[inline, color=red]{#1}}
\newcommand{\suggestions}[1]{\todo[inline, color=yellow]{#1}}
\newcommand{\todos}[1]{\todo[inline, color=cyan]{\textbf{TODO}: #1}}
\newcommand{\outline}[1]{\todo[inline, color=green]{#1}}
%% Utility commands
\newcommand{\hybridaggregate}{\textit{Hybrid Aggregate Computing}}
\newcommand{\round}{\texttt{round}}
\newcommand{\export}{\texttt{export}}
\sloppy
\begin{document}

@@

One of the most challenging fields for Machine Learning application is Collective-Adaptive System (CAS). % Or do we want to use Cyber-Physical Swarm??
 Here, engineers have to handle a massive number of nodes, global-to-local behaviour mapping, and
 highly environment stochasticity.

Historically, researchers favour handling them with self-organisation,
 by which a collective behaviour emerges from continuously local interaction between simple components.

A novel technique to specify self-organising behaviour in a composable and functional manner is Aggregate Computing.
 In particular, this paradigm enables collective program specification as a whole, 
 as it is based on a distributed data structure manipulation called \textit{computational fields}.
 It is successfully used in different scenarios ranging from Swarm Robotics to Smart cities.

Even though it is promising for expressing high-level collective behaviour system-size independent,
 practically it is difficult to make programs work in highly dynamic networks, 
 where nodes have to deal with continuous changes at the environment side.

Therefore, with this work, we stand the first step towards integration between Aggregate Computing and Machine Learning,
 in order to improve adaptivity acting in the low-level specifications but maintaining the high-level program behaviours.

@@

\section{Introduction}
Nowadays, computer engineers have to deal with Collective Adaptive Systems (CASs)
 due to the increasing interconnected computational devices placed in ever-changing environments.
 Examples of those include robot swarms, smart cities and augmented crowds with smart devices.

CAS are notoriously hard~\cite{DBLP:conf/huc/Ferscha15} to design as they are characterised by 
\begin{itemize}
\item a global order that emerges from local node interactions, 
\item decentralised control, and 
\item adaptive behaviours towards environmental changes to pursuit a collective goal.
\end{itemize} 
%
So, over the years, in literature, various approaches have tried to handle this complexity. 
%
Some of them take inspiration from the natural system
 composed of social animals --- like bees, fish, and ants. 
%
Here, a sort of hive mind arises -- 
 also referred to as swarm intelligence --  
 born from local interactions between agents. 
%
These kinds of behaviours are accomplished with self-organisation, 
 namely a global order that emerges from continuous 
 interactions between simple entities.

Initially, researchers had tried to migrate those properties into the computer science world 
 by miming natural collective behaviour bottom-up~\cite{DBLP:journals/connection/Webb02}, 
 namely trying to find the right single-node-behaviour 
 to reach a global behaviour, e.g. flocking, foraging, etc.
%
Though, this mapping isn't easily definable ---
 since we are coping with complex systems.
%
Ideally, we would like to express collective behaviours at the ensemble level,
 abstracting over non-functional aspects such as the network topology, environment uncertainties and 
 system size.

In this direction, Aggregate Computing~\cite{DBLP:journals/computer/BealPV15} is an innovative solution by which 
 developers can express collective self-organising behaviours through a functional program specification.
%
Indeed, an aggregate program (i.e. a program written with Aggregate Computing) consists of the manipulation of \textit{computational fields}, a distributed
 data structure where each point in the spacetime is associated with the result of a computation
 performed by a node.
%
%These operations are defined in \textit{field-calculus}~\cite{DBLP:conf/coordination/AudritoBDV18} --- the root of Aggregate Computing. 
% This calculus describes a set of minimal constructs to express any spatiotemporal computation.

Aggregate Computing captures common abstractions for the collective behaviour specification. 
 These abstractions are encapsulated in the so-called ``building blocks'' through which those various
 applications have been built 
 (such as swarm robotics, crowds engineering~\cite{DBLP:journals/eaai/CasadeiVAPD21}).

Though, building block synthesis concerns different stuff not only related to the behaviour itself
 but also to the "ensemble" specifications. 
 For instance, it is quite difficult to craft building blocks
 that easily adapt w.r.t the node movements rapidity.

Ideally, we would like to continue building our application using these building blocks -- so the high-level specification remains the same -- but using some
 under-the-hood mechanisms that adapt the behaviour according to different environmental changes.

@@

In particular, In this paper,
 we investigate how Aggregate Computing can be extended with Reinforcement Learning (RL) through a case study,
 where we outline the performance improvement of the ``hop count" program --- 
 i.e. an aggregate program that produces a field that stores the distance from a source zone. 
 Furthermore, we show that the learnt policy scales up to different system sizes.
\todos{Outline}

@@

In particular, in literature, we speak about Collective Adaptive Systems~\cite{DBLP:conf/huc/Ferscha15}
 when the number of nodes participating is incredibly high, there is no central authority that coordinate nodes,
 the overall behaviour is reached with local interactions, and the systems expose self-* properties.

Engineering this kind of system is a thriving area, and so various approaches have been proposed.
 First attempts leverage \textit{local-to-global} techniques, where the collective behaviour emerges from local behaviour description, 
 usually inspired by nature~\cite{DBLP:journals/swarm/BrambillaFBD13}. 
%
Though, a natural way to describe this kind of problem is by using a \textit{collective} view of the system.
 This is the idea used by the \textit{global-to-local} techniques~\cite{DBLP:journals/jlap/ViroliBDACP19,DBLP:journals/scp/AlrahmanNL20, DBLP:conf/cbse/BuresGHKKP13}, 
 where the behaviours are designed \textit{top-down}, and then the local behaviour is generated automatically from this ``system-wide'' specification. 
% 
Also ``meet-in-the-middle''~\cite{DBLP:journals/computer/PinciroliB16} approaches exist, where
 a part of the system is still expressed bottom-up (device-centric) whereas a part is expressed top-down (swarm-centric)

In our works, we are focused on Aggregate Computing~\cite{DBLP:journals/computer/BealPV15}, a global-to-local approach where
 the self-organising collective behaviour is expressed in terms of manipulation of \textit{computational fields}.

The latter could be seen as the extension of electromagnetic fields to Computer Science since it consists of a 
 map from any devices to their computational value (i.e. the result of a computation).

The computational field manipulation is ruled by \textit{field calculus}, which define the constructs needed 
 to express spatio-temporal computation~\cite{DBLP:conf/coordination/AudritoBDV18}.

The novelty of Aggregate Computing consists in its composable (functional-like) system-size independent program definition: 
using basic constructs we identify the main patterns (\textit{building-blocks}) that could be reusable at the application level.

By opportunistically crafting these building-blocks it is possible to verify prominent properties such as eventual consistency~\cite{DBLP:journals/taas/BealVPD17} 
 (i.e. the ensemble behaviour is independent of the underlying network details)
 and self-stabilisation~\cite{DBLP:journals/tomacs/ViroliABDP18} (i.e. the ability of a system to recover from arbitrary changes).

Aggregate Computing is studied in various simulated scenarios ranging from smart cities to crowds, 
 thanks to Alchemist~\cite{DBLP:journals/jos/PianiniMV13} (a meta-simulator for pervasive systems) and to programming languages and toolchains that support 
 this paradigm, such as ScaFi~\cite{DBLP:conf/ecoop/CasadeiV16}, FCCP~\cite{DBLP:conf/acsos/Audrito20} and Protelis~\cite{PianiniSAC2015}.

In the next sections, we show the field calculus operators that are essential to understand the case study presented,
 and then we underline how each node computes in order to follow the aggregate program specification.
\subsection{Field Calculus operators}
The main constructs that capture the essential aspects for programming self-organising systems are:
\begin{itemize}
  \item \textit{Stateful field evolution} expression \mintinline[escapeinside=||]{scala}{|{\color{teal}rep}|(|$e_1$|) {|($x$) => $e_2$|}} describes a field evolving in time. 
  $e_1$ is the initial field value and the function $(x) => e_2$ define how the field changes at each execution.
  \item \textit{Neighbour interaction} expression \mintinline[escapeinside=||]{scala}{|\color{teal}nbr|{|$e$|}} a view
  of the field values in the surroundings of each device where neighbours are
  mapped to their evaluations of $e$. A series of \mintinline{scala}{*hood} operators leverage this map to produce relevant result. 
  For instance, \mintinline{scala}{minHood} return the minimum value 
  within the neighbourhood.
  \item \textit{Domain partitioning} expression \mintinline[escapeinside=||]{scala}{|\color{teal}if|(|$e_0$|){|$e_1$|}{|$e_2$|}} splits the computational
  field into two non-communicating zones hosting isolated subcomputations:
  $e_1$ where $e_0$ is true, and $e_2$ where $e_0$ is false.
\end{itemize}

For a more in-depth explanation of these operators, please refer to~\cite{DBLP:journals/computer/BealPV15}.
\subsection{Computational Model}
An aggregate program execution consists of a continuously node-point execution of a \round{}, 
 that is the atomical part of aggregate computing.
%
Each node can interact only with-in its neighbourhoods sharing the \export{}, 
 namely the result produced by the evaluation of a \round{}.

The main round steps consist of:
\begin{enumerate}
  \item execution context creation: nodes create the execution context collecting information from the sensors, 
  the neighbourhood messages, and the old program output;
  \item program evaluation: the system-wide aggregate program is evaluated against the local context created by the nodes. 
  This produces the \export{};
  \item export sharing: the export created will be shared within the current neighbourhoods perceived by the nodes;
  \item actuations: giving the export produced, nodes may perform some actuation (e.g. motor activations, steering, ...).
\end{enumerate}

@@

In this section,
 we analyse the characteristics of the aggregate computing paradigm,
 map these to relevant machine learning contributions in literature,
 and pointing out different learning perspectives about the integration of Aggregate Computing with Machine Learning

\paragraph{Multi-and many-agent system}
%
An aggregate system is a multi-agent system
 and, often, a \emph{many}-agent system
 where a large number (hundreds or more)
 of autonomous entities are programmed to achieve 
 some collective behaviour through \emph{repeated} 
 sensing, computation, communication, and actuation steps.
%
Due to the high stochasticity of the environment,
 it is almost impossible to know and
 program the optimal behaviour for all agents in advance.
 This uncertainty results in the need of creating intelligent agents
 so that they can learn the optimal behaviour and adapt to environmental changes.
%
In recent decades there has been an emerging trend in the use of Reinforcement Learning 
 in multi-agent settings (called Multi-Agent Reinforcement Learning), as a powerful, robust and adaptive learning paradigm.
 Progress has been considerable and a wide range of algorithms are now available.

Multi-Agent Reinforcement Learning is a conjunction of Game theory and Reinforcement Learning, 
 and there are several (even orthogonal) viewpoints on which researchers have been focused.

First attempts from RL viewpoint, 
 goes towards a so-called \textit{independent learning}~\cite{DBLP:journals/tsmc/BusoniuBS08} approach, 
 where each agent learn locally against the whole environment~\cite{DBLP:conf/icml/Tan93}.

Examples of these are Hysteretic Q-Learning (used to reduce non-stationarity)~\cite{hysteretic-q}, 
 Lenient Learning~\cite{DBLP:journals/jmlr/WeiL16}, and Frequency Maximum Q-value~\cite{DBLP:conf/atal/KaisersT10}.
%
The pro of these approaches is that their complexity does not scale up with the number of agents, 
 but, unfortunately, the learning process is extremely non-stationary and unstable.
% 
Furthermore, they are heuristic and do not exist any convergence proof (even if, in practice, they often reach a good policy).

The game theory viewpoint instead has tried to find techniques that achieve equilibrium conditions such as Nash-equilibrium of Pareto optimality.
 The first works towards this direction are Nash-Q-Learning~\cite{nash-q}, Friend-or-Foe Q-learning~\cite{DBLP:conf/icml/Littman01}, Minimax-Q~\cite{DBLP:conf/icml/Littman94}.
The problem is that it is used with few agents and scale hardly with the application complexity.
%
Finally, current emerging trends tend to leverage a so-called centralised training and decentralised execution (CTDE) approach, by which 
 agents should leverage the system-wide knowledge at training time but at runtime, they act independently~\cite{DBLP:journals/aamas/Hernandez-LealK19}.
 Most promising approaches in this direction are MAPPO~\cite{DBLP:journals/corr/abs-2103-01955}, COMA~\cite{DBLP:journals/corr/FoersterFANW17} and MADDPG~\cite{DBLP:conf/nips/LoweWTHAM17}.

The main problem with most of the solutions available in the literature is that they consider a small number of agents (or at least test them on small games).
 In Aggregate Computing instead, we need to find a technique that does not depend on the system node count.

\paragraph{Neighbour-based or indirect (environment-mediated) interaction}
%
In aggregate systems, a device can directly interact only with neighbours.
%
Data flows may be implemented
 to support indirect communication across multiple hops;
 however, information from devices far away tends to be obsolete.
%
The environment can also be used, via stigmergy~\cite{DBLP:journals/scp/NicolaSI20},
 for indirect communication;
 however, a device can typically only access 
  to a very local portion of the environment.
%
In other words, the system state is only \emph{partially observable}.

@@

A way to handle this kind of rewards in multi-agent settings is \textit{difference rewards}~\cite{DBLP:conf/atal/DevlinYKT14}. It is used to capture agents contributions to the system's global performance.
\paragraph{Transient phase}
%
When an aggregate behaviour is carried out,
 we often focus on the stabilised outcomes,
 tolerating outcomes that are generated
 during the transient phase.
%
However, intermediate results are often important,
 and invariants should hold throughout the entire computation.

@@

\subsection{Integration perspectives}
Here we outline the different perspectives about Aggregate Computing with learning, 
 to clarify the current and the future research path.
%
\paragraph{Application} Learning can be useful at different levels in our framework.
 First of all, learning can be applied to the building blocks definition.
% 
Indeed, most of our results and applications are based on them.
%
The building block improvement may consist either of learning a correction factor or by entirely learning the behaviour.

On the other side, learning can be useful also for framework related stuff,
 mainly aiming at "non-functional" aspects. 
 Trust mechanisms, evaluation frequency, energy consumption manager, package storage are all the things that could take advantage of using Machine Learning algorithms.

\paragraph{Learning problem} 
The learning problem in multi-agent settings can be posed as \textit{Independent} and \textit{Team} learning. 
In the foster, each agent perform the learning process and while in the latter,
 the learning process is performed only by one agent and then shared with the collective (in a typical CDTE setting).
% 
Team learning is usually used in the ``swarm'' system, due to the homogeneous behaviour 
 --- i.e. each agent performs the same program. 
 Here, a common reward signal exists and evaluate the overall collective behaviour.
% 
In Aggregate Computing, each agent follows the \textit{same} aggregate program shared within the entire system, so Team learning can 
 be applied seamlessly.

On the other hand, Independent Learning is usually applied in Collective
 (Self-)Adaptive System \cite{csas-and-marl}, a field near to Aggregate Computing.

In particular, Independent Reinforcement Learning reach good results in this kind of system, 
 even if no theorem exists as in the single-agent context.
%
With Independent learning though is still possible to reach a global collective behaviour convergence. 
 Indeed using a common reward signal, the collective may reach the same behaviour through emergence \cite{iima2008swarm, nguyen2018swarm}.

\paragraph{Learning technique}
Aside from what kind of problem we set up, we can then use different Machine Learning techniques,
 such as Reinforcement Learning or Supervised Learning.
%
Supervised learning is rare in CSAS -- or in the ``swarm" like system in general.
%
Novel works~\cite{DBLP:conf/corl/TolstayaGPP0R19} leverage Graph Neural network to learn how agents should communicate, 
 supposing to know the right behaviour (i.e. produced via global vision simulation or by emulation of natural phenomena).

In Aggregate Computing, Supervised Learning can be applied when we want to build up new building blocks from zero. 
% 
Indeed, we eventually know the right state of the system, but we do not know how the system can reach it.

RL is predominantly used in CAS, 
 due to its flexibility and thanks to the plethora of approaches used in this scenario~\cite{csas-and-marl}.

In our case, RL can be used both at the building block level and the framework level.
%
For the foster, the simplest scenario can be that agents learn a correction factor to improve a long-term reward signal.

\paragraph{Learning time}
Finally, learning can be performed \textit{offline}, \textit{online} or \textit{offline an then online} (\textit{mixed time}).
% 
The foster is the common scenario in multi-agent settings.
 Here indeed, we can leverage simulation to know the right computational field and then use one of the novel CTDE techniques.

Mixed-time learning can be a mid-level complexity direction instead.
 Here, simulation can always be exploited to reach a good behaviour, but at runtime, 
 the behaviour can be adapted based on the new environmental situations.

@@

\section{Aggregate Computing with Reinforcement}
\begin{figure}
  \centering
  \includegraphics[width=\linewidth]{img/algorithm-learning.pdf}
  \caption{This figure shows the main steps of RL in Aggregate Computing.
  Each agent starts with his copy of a Q table. 
  Then, during the simulation, they refine them Q tables changing q values according to their experience (different colours mean different values). 
  In the end, a new $Q_{max}$ table will be created aggregating each agent table. 
  Finally, a new episode starts using this $Q_{max}$ as a starting Q table.
  }
  \label{fig:aggregate-q-learning}
\end{figure}
As first experiments, we leverage RL to improve building blocks. 
% 
In this section, we briefly describe how Reinforcement Learning can be injected into the computation loop,
 and what Reinforcement Learning we used with Aggregate Computing.

From a local view, program execution can be seen as:
$$
f : \textit{Context} \rightarrow \textit{Output}
$$
Where $\textit{Context}$ contains:
\begin{itemize}
  \item $\sigma(name) \rightarrow \textit{data}$: a function from sensor name to the perceived value;
  \item $\phi(\textit{ID}) \rightarrow \textit{Export}$: a function from id to neighbourhoods export data.
\end{itemize}
The computational field $\theta$ contains all output produced by the system.
